Three Algorithms for Merging Hierarchical Navigable Graphs

### 2025-04-30 paper/main.tex
add section Graph Merging as an Iterative Process

\title{Three Algorithms for Merging (HNSW) Navigable Graphs}
\author[1]{Alexander Ponomarenko}
\date{}

@@

% TODO для экспериментов. 
% 1. Оценить сколько вычислений функции расстояния нужно, для того чтобы просто добавить все элементы одного графа в другой просто использу операцию вставки. Оценить у конечного графа recall/distance computation number 
% 2. Провести эксперименты для разных размерностей и размеров
% 3. Следить за тем сколько операций "полного поиска было сделано" и сколько операций "local search".
% 4. Нарисовать графики хактеристик смердженого графа recall/distance computation number в зависимости от параметра merge-алгоритма и числа ресурсов потраченных для формирования объединёного графа
% 5. Сделать все теже эксперементы, но используя в качестве NeighborhoodConstruction k-nn

\section{Introduction}
Navigable small-world graphs have emerged as a cornerstone for efficient approximate nearest neighbor (ANN) search in high-dimensional spaces. This problem is crucial in fields such as recommendation systems, image retrieval, and natural language processing. Algorithms like Hierarchical Navigable Small World (HNSW) \cite{hnsw} and its predecessor Navigable Small World (NSW) \cite{nsw2011,nsw2012,nsw2014} provide robust and scalable frameworks for organizing data into graph structures that support fast similarity searches.

One of the key challenges in maintaining these graph-based indices is efficiently combining data from multiple sources. As datasets grow and evolve, it becomes necessary to merge separate graph structures while preserving their navigability properties. This is particularly important in distributed systems where indices are built independently and later consolidated, or when integrating new batches of data with existing indices.

While HNSW supports dynamic insertions, merging entire graphs as coherent structures presents a different challenge. The quality of the merged graph directly impacts search performance, and naive approaches may lead to suboptimal neighborhood connections or excessive computational costs during the merge operation.

In this work, we propose three novel algorithms—\textbf{Naive Merge}, \textbf{Intra Graph Traversal Merge (IGTM)}, and \textbf{Cross Graph Traversal Merge (CGTM)}—for efficiently merging navigable graphs. While our methods primarily target HNSW, they are generalizable to other graph-based data structures such as NSW \cite{nsw2011} and NSG \cite{NSG}. These algorithms preserve graph navigability and maintain the structural integrity necessary for fast ANN search while offering different trade-offs between merge accuracy and computational efficiency.

>>>>

\title{Three Algorithms for Merging Hierarchical Navigable Graphs}
\author[1]{Alexander Ponomarenko}
\date{}

@@

% TODO для экспериментов. 
% 2. Провести эксперименты для разных размерностей и размеров
% 3. Следить за тем сколько операций "полного поиска было сделано" и сколько операций "local search".
% 5. Сделать все теже эксперементы, но используя в качестве NeighborhoodConstruction k-nn

\section{Introduction}

Navigable small-world graphs have emerged as a cornerstone for efficient approximate nearest neighbor (ANN) search in high-dimensional spaces. This problem is crucial in fields such as recommendation systems, image retrieval, and natural language processing. Algorithms like Hierarchical Navigable Small World (HNSW) \cite{hnsw} and its predecessor Navigable Small World (NSW) \cite{nsw2011,nsw2012,nsw2014} provide robust and scalable frameworks for organizing data into graph structures that support fast similarity searches.

\subsection{Nearest Neighbor Search Problem}

Formally, the nearest neighbor search problem can be defined as follows: Given a set of points $P = \{p_1, p_2, ..., p_n\}$ in a metric space $(X, \rho)$, where $\rho$ is a distance function, and a query point $q \in X$, find the point $p^* \in P$ such that:

$$p^* = \underset{p \in P}{\mathrm{argmin}} \rho(q, p)$$

In high-dimensional spaces, this problem becomes computationally challenging due to the ``curse of dimensionality.'' Exact solutions often degrade to linear scan as dimensionality increases, making them impractical for large-scale applications. This has led to the development of approximate nearest neighbor (ANN) search methods, which trade perfect accuracy for substantial improvements in computational efficiency.

\subsection{Navigable Graphs for ANN Search}

Three main approaches have been developed for tackling the ANN search problem:

\begin{enumerate}
    \item \textbf{Space-partitioning trees}: These methods divide the vector space into hierarchical regions, enabling logarithmic search complexity in low dimensions. Notable examples include KD-trees \cite{Bentley1975}, R-trees \cite{Guttman1984}, VP-trees \cite{Yianilos1993}, M-trees \cite{Ciaccia1997}, and Cover trees \cite{Beygelzimer2006}. A comprehensive overview of these approaches for metric spaces is provided by Zezula et al. \cite{Zezula2006}.

    \item \textbf{Mapping-based methods}: These techniques transform the original data into representations that allow for efficient search. This category includes Locality-Sensitive Hashing (LSH) \cite{Indyk1998,Gionis1999,Datar2004,Andoni2006,Andoni2015} and Product Quantization approaches \cite{Jegou2011,Ge2013,Norouzi2013}, which encode vectors into compact codes that approximate distances between points.

    \item \textbf{Navigable graph-based methods}: The most recent and currently state-of-the-art approach that constructs graph structures where vertices represent data points and edges connect similar points. These graphs are designed to allow ``greedy-like'' algorithms to perform directed traversal through the data space.
\end{enumerate}

A navigable graph is a graph $G = (V, E)$ where $V$ corresponds to the set of data points $P$, and the edge set $E$ is constructed to enable efficient navigation toward any target point using local information only. The key insight is that with properly constructed edges, a simple greedy search starting from any node can efficiently converge to the vicinity of a query point by following edges that minimize the distance to the query at each step.

The effectiveness of navigable graphs relies on the ``small-world'' property—a combination of short-range connections that create clusters of similar points and long-range connections that allow quick transitions between distant regions of the space. This structure enables logarithmic or sub-logarithmic search time complexity in practice, even though the theoretical worst-case bounds may be higher.

\subsection{Hierarchical Navigable Small World (HNSW)}

HNSW \cite{hnsw} extends the navigable small-world concept by organizing points into a hierarchy of layers, each containing a progressively smaller subset of the data points. The bottommost layer contains all points, while higher layers form a sparse representation of the data space. This hierarchical structure allows for a coarse-to-fine search strategy: starting at the highest layer with the entry point, the search first navigates to the approximate region of the target at a coarse level, then descends through layers while refining the approximation.

HNSW supports two primary operations:
\begin{enumerate}
    \item \textbf{Search}: Efficient retrieval of approximate nearest neighbors for a given query point
    \item \textbf{Insertion}: Dynamic addition of new data points to the existing structure while maintaining navigability properties
\end{enumerate}

The insertion operation constructs appropriate connections for a new point across all layers it belongs to, ensuring that both short-range and long-range connections are established to maintain the small-world property.

\subsection{Graph Merging as an Iterative Process}

One of the key challenges in maintaining these graph-based indices is efficiently combining data from multiple sources. While HNSW supports dynamic insertions, merging entire graphs as coherent structures presents a different challenge. The quality of the merged graph directly impacts search performance, and naive approaches may lead to suboptimal neighborhood connections or excessive computational costs during the merge operation.

It is worth noting that graph merging can be used as an analog to compaction operations in database systems. In graph-based indices, deletion operations are often implemented by simply marking vertices as deleted without actually removing them, to avoid disrupting the graph's connectivity. This approach prevents the computationally expensive task of rebuilding neighborhoods for affected vertices. Moreover, a true deletion might trigger cascading neighborhood reconstructions that could potentially affect the entire graph, which is unacceptable from a performance perspective. Graph merging provides a way to periodically compact the structure by creating a fresh graph that excludes deleted vertices.

The merge operation can be conceptualized as an iterative process consisting of four key steps:

\begin{enumerate}
    \item \textbf{Vertex Selection}: Choosing a vertex $v^*$ for which we aim to construct a neighborhood in the merged graph, and determining which original graph it belongs to.

    \item \textbf{Candidate Collection}: Gathering potential neighbors from both input graphs. This involves searching within each graph structure to find vertices that should be considered for inclusion in $v^*$'s neighborhood.

    \item \textbf{Neighborhood Construction}: Applying a specified strategy (e.g., k-nearest neighbors or relative neighborhood graph) to select the final set of neighbors for $v^*$ from the candidates.

    \item \textbf{Information Propagation}: Deciding what information from the current iteration should be preserved for the next iteration to optimize the process, then proceeding to step 1 with a new vertex.
\end{enumerate}

By carefully designing each of these steps, we can create merge algorithms that balance computational efficiency with the quality of the resulting graph structure.

In this work, we propose three novel algorithms based on this iterative framework—\textbf{Naive Graph Merge (NGM)}, \textbf{Intra Graph Traversal Merge (IGTM)}, and \textbf{Cross Graph Traversal Merge (CGTM)}—for efficiently merging navigable graphs. These algorithms differ primarily in how they implement each step of the iterative process. For instance, NGM and IGTM select the next vertex for neighborhood construction only from the same graph as the current vertex, promoting locality within original graphs. In contrast, CGTM implements the idea that the next vertex for neighborhood construction can be selected from any input graph, enabling more thorough exploration of the combined space. While our methods primarily target HNSW, they are generalizable to other graph-based data structures such as NSW \cite{nsw2011} and NSG \cite{NSG}. These algorithms preserve graph navigability and maintain the structural integrity necessary for fast ANN search while offering different trade-offs between merge accuracy and computational efficiency.

The rest of this paper is organized as follows: Section 2 describes the search algorithms used in navigable small-world graphs. Section 3 presents neighborhood construction strategies that are essential for both graph building and merging. Section 4 introduces our three graph merging algorithms in detail. Section 5 presents experimental results comparing the proposed methods across standard benchmarks. Finally, Section 6 concludes the paper and discusses directions for future research.
